A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-05-17 JNABsub20230515/ComparingNLStools.Rmd
add submission may 23

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-3-2"
output:
  pdf_document:
    keep_tex: yes
    toc: no
  word_document:
    toc: no
bibliography: ./ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## require(bookdown) # language engine to display text - does not seem necessary
```

## Abstract

Our Google Summer of Code project "Improvements to `nls()`" considered the features and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, including our `nlsr`, provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in
Fernandez Torvisco et al. (2018) for fitting separable nonlinear regression curves. The 
special class of problem for which this package is intended is an important and 
difficult one. No vignette is provided, unfortunately.

@@

with suggested addition of small perturbations to the data. This admits 
`nls()` could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Fortunately, a small adjustment
to the "termination test" for the **program**, rather than for the
"convergence" of the underlying **algorithm**, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see @BatesWatts81). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
additional parameter `scaleOffset` with a default value of zero. Setting it to a 
small number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**, and it has been in `nlsr`
since it was introduced. The default value 
gives the legacy behaviour. This improvement has been in the R distributed code 
since version 4.1.0.

Additional termination tests can be used. `nlsr` has a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, usually minor, but sometimes annoying with
ill-conditioned problems.

@@

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls`, and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods which we
have not investigated.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants. In general, the Levenberg-Marquardt stabilization 
is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors.

@@

`na.action` is an argument to the `nls()` function, but it does not appear 
obviously in the source code, often being handled behind the scenes after 
referencing the option `na.action`. This feature also changes the data 
supplied to our nonlinear least squares solver. 
A useful, but possibly dated, description is given in:
*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
The typical default action, which can be seen by using the command
`getOption("na.action")`
is `na.omit`. This option removes from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for solver computations, but keeps the rows with 
NA elements so that predictions are in the correct row position. We recommend that 
workers actually test output to verify the behaviour is as wanted.
See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
As with `subset`, our concern with `na.action` is that users may be unaware of the
effects of an option they may not even be aware has been set. Should `na.fail` be the default?

@@

The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
as does, for example, `result$m$resid()`. The function `nlsModel()`, which we 
have had to extract from the base R code and explicitly `source()` because it 
is not exported to the working namespace,
allows us to compute residuals for particular coefficient sets.

```{r nlswtx, echo=TRUE}
wts <- 0.5^tt # simple weights
frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
Asym<-1; xmid<-1; scal<-1
nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
attr(rnowt, "gradient")<-NULL; rnowt
usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
attr(rusewt,"gradient")<-NULL; rusewt
source("nlsModel.R")
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0
nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
rn<-nmod$resid()
attr(rn,"gradient")<-NULL; rn
```

@@

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).
In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

>>>>

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
date: "2023-5-16"
abstract: >
  Our Google Summer of Code project "Improvements to `nls()`" considered the 
  features and limitations of the R function `nls()` with the aim 
  of improving and rationalizing R tools for nonlinear regression.
  The rich features of `nls()` are weakened by several deficiencies and
  inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
  Further considerations are the usability and  maintainability of
  the code base that provides the functionality `nls()` claims to offer.
  Various packages, including our `nlsr`, provide alternative capabilities.
  We consider the differences in goals, approaches, and features of
  different tools for nonlinear least squares modeling in R. Discussion of 
  these matters is relevant to improving R generally as well as its 
  nonlinear estimation tools.

draft: true
author:
  # see ?rjournal_article for more information
    - name: John C. Nash
      affiliation: retired professor, University of Ottawa
      address:
      - Telfer School of Management
      - Ottawa ON Canada K1N 6N5
      orcid: 0000-0002-2762-8039
      email: profjcnash@gmail.com
    - name: Arkajyoti Bhattacharjee
      affiliation: Indian Institute of Technology
      address:
      - Department of Mathematics and Statistics
      - Kanpur
      email: arkastat98@gmail.com
type: package
output:
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: ./ComparingNLStools.bib

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

@@

`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in
@Torvisco2018 for fitting separable nonlinear regression curves. The 
special class of problem for which this package is intended is an important and 
difficult one. No vignette is provided, unfortunately.

@@

with suggested addition of small perturbations to the data. This admits 
`nls()` could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Fortunately, a small adjustment
to the "termination test" for the **program**, rather than for the
"convergence" of the underlying **algorithm**, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see @BatesWatts81). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
additional parameter `scaleOffset` with a default value of zero. Setting it to a 
small number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**, and it has been in `nlsr`
since it was introduced. The default value 
gives the legacy behavior. This improvement has been in the R distributed code 
since version 4.1.0.

Additional termination tests can be used. `nlsr` has a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
%% ^ is picked up in rjtools check

@@

`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behavior, usually minor, but sometimes annoying with
ill-conditioned problems.

@@

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls`, and `minpack.lm` use flavors of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods which we
have not investigated.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants. In general, the Levenberg-Marquardt stabilization 
is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors.

@@

`na.action` is an argument to the `nls()` function, but it does not appear 
obviously in the source code, often being handled behind the scenes after 
referencing the option `na.action`. This feature also changes the data 
supplied to our nonlinear least squares solver. 
A useful, but possibly dated, description is given in:
*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
The typical default action, which can be seen by using the command
`getOption("na.action")`
is `na.omit`. This option removes from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for solver computations, but keeps the rows with 
NA elements so that predictions are in the correct row position. We recommend that 
workers actually test output to verify the behavior is as wanted.
See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
As with `subset`, our concern with `na.action` is that users may be unaware of the
effects of an option they may not even be aware has been set. Should `na.fail` be the default?

@@

The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
as does, for example, `result$m$resid()`. The function `nlsModel()` 
allows us to compute residuals for particular coefficient sets.
We have had to extract from the base R code and include it via a code chunk 
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly `source()` this code.

```{r nlsmodelsource, echo=FALSE}
#  From file src/library/stats/R/nls.R
#  Part of the R package, https://www.R-project.org
#
#  Copyright (C) 2000-2020 The R Core Team
#  Copyright (C) 1999-1999 Saikat DebRoy, Douglas M. Bates, Jose C. Pinheiro
#
#  The nlsModel() function extracted here.

nlsModel <- function(form, data, start, wts, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
{
    ## thisEnv <- environment() # shared by all functions in the 'm' list; variable no longer needed
    env <- new.env(hash = TRUE, parent = environment(form))
    for(i in names(data)) env[[i]] <- data[[i]]
    ind <- as.list(start)
    parLength <- 0L
    for(i in names(ind) ) {
        temp <- start[[i]]
        storage.mode(temp) <- "double"
        env[[i]] <- temp
        ind[[i]] <- parLength + seq_along(temp)
        parLength <- parLength + length(temp)
    }
    getPars.noVarying <- function() unlist(mget(names(ind), env))
    getPars <- getPars.noVarying
    internalPars <- getPars()

    if(!is.null(upper)) upper <- rep_len(upper, parLength)
    useParams <- rep_len(TRUE, parLength)
    lhs <- eval(form[[2L]], envir = env)
    rhs <- eval(form[[3L]], envir = env)
    .swts <- if(!missing(wts) && length(wts))
        sqrt(wts) else rep_len(1, length(rhs))
    env$.swts <- .swts
    resid <- .swts * (lhs - rhs)
    dev <- sum(resid^2)
    if(is.null(attr(rhs, "gradient"))) {
        getRHS.noVarying <- function() {
            if(is.null(upper)) # always for "default"
                numericDeriv(form[[3L]], names(ind), env, central = nDcentral)
            else # possibly with "port"
                numericDeriv(form[[3L]], names(ind), env,
                             dir = ## ifelse(internalPars < upper, 1, -1)
                                 -1 + 2*(internalPars < upper), central = nDcentral)
        ## 210701JN: Above line should check for nDcentral=FALSE when approaching
        ## upper bound, as central diff is same whether dir is +1 or -1. ???
        }
        getRHS <- getRHS.noVarying
        rhs <- getRHS()
    } else {
        getRHS.noVarying <- function() eval(form[[3L]], envir = env)
        getRHS <- getRHS.noVarying
    }
    dimGrad <- dim(attr(rhs, "gradient"))
    marg <- length(dimGrad)
    if(marg > 0L) {
        gradSetArgs <- vector("list", marg + 1L)
        for(i in 2L:marg)
            gradSetArgs[[i]] <- rep_len(TRUE, dimGrad[i-1L])
        useParams <- rep_len(TRUE, dimGrad[marg])
    } else {
        gradSetArgs <- vector("list", 2L)
        useParams <- rep_len(TRUE, length(attr(rhs, "gradient")))
    }
    npar <- length(useParams)
    gradSetArgs[[1L]] <- (~attr(ans, "gradient"))[[2L]]
    gradCall <-
        switch(length(gradSetArgs) - 1L,
               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], drop = FALSE),
               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], gradSetArgs[[2L]],
                    drop = FALSE),
               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], gradSetArgs[[2L]],
                    gradSetArgs[[3L]], drop = FALSE),
               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], gradSetArgs[[2L]],
                    gradSetArgs[[3L]], gradSetArgs[[4L]], drop = FALSE))
    getRHS.varying <- function()
    {
        ans <- getRHS.noVarying()
        attr(ans, "gradient") <- eval(gradCall)
        ans
    }
    if(length(gr <- attr(rhs, "gradient")) == 1L)
		    attr(rhs, "gradient") <- gr <- as.vector(gr)
    QR <- qr(.swts * gr)
    qrDim <- min(dim(QR$qr))
    if(QR$rank < qrDim)
        stop("singular gradient matrix at initial parameter estimates")

    getPars.varying <- function() unlist(mget(names(ind), env))[useParams]
    setPars.noVarying <- function(newPars)
    {
        internalPars <<- newPars # envir = thisEnv
        for(i in names(ind))
            env[[i]] <- unname(newPars[ ind[[i]] ])
    }
    setPars.varying <- function(newPars)
    {
        internalPars[useParams] <<- newPars
        for(i in names(ind))
            env[[i]] <- unname(internalPars[ ind[[i]] ])
    }
    setPars <- setPars.noVarying

    if(scaleOffset) scaleOffset <- (length(resid)-npar) * scaleOffset^2
    convCrit <- function() {
        if(npar == 0) return(0)
        rr <- qr.qty(QR, c(resid)) # rotated residual vector
        sqrt( sum(rr[1L:npar]^2) / (scaleOffset + sum(rr[-(1L:npar)]^2)))
    }

    on.exit(remove(i, data, parLength, start, temp, m, gr,
                   marg, dimGrad, qrDim, gradSetArgs))
    ## must use weighted resid for use with "port" algorithm.
    m <-
	list(resid = function() resid,
	     fitted = function() rhs,
	     formula = function() form,
	     deviance = function() dev, ##?? PROBLEM -- does not get updated
	     lhs = function() lhs,
	     gradient = function() .swts * attr(rhs, "gradient"),
	     conv = function() convCrit(),
	     incr = function() qr.coef(QR, resid),
	     setVarying = function(vary = rep_len(TRUE, np)) {
                 np <- length(useParams)
		 useParams <<- useP <-
                     if(is.character(vary)) {
                         temp <- logical(np)
                         temp[unlist(ind[vary])] <- TRUE
                         temp
                     } else if(is.logical(vary) && length(vary) != np)
                         stop("setVarying : 'vary' length must match length of parameters")
                     else
                         vary # envir = thisEnv
		 gradCall[[length(gradCall) - 1L]] <<- useP
		 if(all(useP)) {
		     setPars <<- setPars.noVarying
		     getPars <<- getPars.noVarying
		     getRHS  <<-  getRHS.noVarying
		     npar    <<- length(useP)
		 } else {
		     setPars <<- setPars.varying
		     getPars <<- getPars.varying
		     getRHS  <<-  getRHS.varying
		     npar    <<- sum(useP)
		 }
	     },
	     setPars = function(newPars) {
		 setPars(newPars)
		 resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
		 dev   <<- sum(resid^2) # envir = thisEnv
		 if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
		 QR <<- qr(.swts * gr) # envir = thisEnv
		 (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
	     },
	     getPars = function() getPars(),
	     getAllPars = function() getPars(),
	     getEnv = function() env,
	     trace = function() {
		 d <- getOption("digits")
		 cat(sprintf("%-*s (%.2e): par = (%s)\n", d+4L+2L*(scaleOffset > 0),
			     formatC(dev, digits=d, flag="#"),
			     convCrit(),
			     paste(vapply(getPars(), format, ""), collapse=" ")))
	     },
	     Rmat = function() qr.R(QR),
	     predict = function(newdata = list(), qr = FALSE)
                 eval(form[[3L]], as.list(newdata), env)
	     )
    class(m) <- "nlsModel"
    m
}
```

```{r nlswtx, echo=TRUE}
wts <- 0.5^tt # simple weights
frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
Asym<-1; xmid<-1; scal<-1
nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
attr(rnowt, "gradient")<-NULL; rnowt
usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
attr(rusewt,"gradient")<-NULL; rusewt
## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0
nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
rn<-nmod$resid()
attr(rn,"gradient")<-NULL; rn
```

@@

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).
In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarization of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.
